---
title: "Quick NHANES Video"
author: "Thomas E. Love"
date: "`r Sys.Date()`"
output: 
    html_document:
        number_sections: TRUE
        toc: TRUE
        code_folding: show
        toc_float: TRUE
---

# Setup

```{r setup}
knitr::opts_chunk$set(comment = NA)

library(nhanesA)
library(janitor)
library(magrittr)
library(tidyverse)
```

# Ingest the NHANES data

Three NHANES databases.

- Demographics (DEMO_J)
- Lab Data (Complete Blood Count with 5-Part Differential) (CBC_J)
- Questionnaire Data (Current Health Status) (HSQ_J)

```{r}
demo_raw <- nhanes('DEMO_J') %>% tibble()
cbc_raw <- nhanes('CBC_J') %>% tibble()
hsq_raw <- nhanes('HSQ_J') %>% tibble()
```

From Demographics, I want the following variables.

- SEQN
- RIDSTATR
- RIDAGEYR
- DMDBORN4

```{r}
demo1 <- demo_raw %>% 
    select(SEQN, RIDSTATR, RIDAGEYR, DMDBORN4) %>%
    filter(complete.cases(.))

dim(demo1)
```

From CBC data, I want:

- SEQN
- LBXWBCSI

```{r}
cbc1 <- cbc_raw %>%
    select(SEQN, LBXWBCSI) %>%
    filter(complete.cases(.))

dim(cbc1)
```

From the HSQ data, I want:

- SEQN
- HSD010

```{r}
hsq1 <- hsq_raw %>%
    select(SEQN, HSD010) %>%
    filter(complete.cases(.))

dim(hsq1)
```

# Merge these data sets

```{r}
temp_new <- left_join(demo1, cbc1, by = "SEQN")

dim(temp_new)
```

```{r}
merged_data <- left_join(temp_new, hsq1, by = "SEQN") %>%
    filter(complete.cases(.))

dim(merged_data)
```

# Clean up these data

```{r}
Hmisc::describe(merged_data)
```

```{r}
merged_data %>% tabyl(DMDBORN4)
```

In DMDBORN4, the codes mean the following:

- 1 means in the US
- 2 means outside the US
- 77 means Refused and should be treated as NA
- 99 means Don't Know and should be treated as NA

We want to change the 77 and 99 to NA, and then relabel 1 and 2 to be more meaningful.

```{r}
dim(merged_data)

mydata <- merged_data %>%
    mutate(DMDBORN4 = na_if(DMDBORN4, "77"),
           DMDBORN4 = na_if(DMDBORN4, "99")) %>%
    mutate(DMDBORN4 = 
               fct_recode(factor(DMDBORN4),
                          "US" = "1",
                          "non-US" = "2")) %>%
    filter(complete.cases(.)) 

dim(mydata)

mydata %>% tabyl(DMDBORN4)
```
